Deep Neural Architectures for Discourse Segmentation in E-Mail Based Behavioral Interventions

### 2018-11-14 paper-amia-summit-2019/manuscript_marked.tex
duplicate manuscript for highlighing the changes

\documentclass{amia}
\usepackage{graphicx}
\usepackage[labelfont=bf]{caption}
\usepackage[superscript,nomove]{cite}
\usepackage{color}
\usepackage{multirow}
\renewcommand*{\thefootnote}{\alph{footnote}}

@@

\section*{Introduction}
The emergence of e-Health technologies opened up new ways to deliver a variety of behavioral interventions to any demographic group of patients in any geographical location. Motivational interviewing (MI), an evidence-based communication technique to increase intrinsic motivation and self-efficacy for behavior change,\cite{miller2012motivational,miller2009toward} is one type of these interventions. MI sessions are generally aimed at eliciting ``change talk'', or statements of intrinsic motivation about patients' own desire, ability, reasons, need for and commitment to behavior change, which have been established by previous research\cite{apodaca2009mechanisms} as a reliable mediator of health behavior change. However, communication science approaches to understanding the efficacy of MI are inherently limited by traditional qualitative coding methods.

@@

The goal of this study is to assess the effectiveness of deep learning methods for the task of automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of qualitative analysis of this type of clinical communications. Specifically, for this study, we utilized the data from MENU GenY (Making Effective Nutrition Choices for Generation Y) \cite{alexander2017motivations}, a recent technology-based public health intervention that relies on personalized e-Coaching to encourage increased fruit and vegetable intake among young adults, aged 21--30. The goal of MENU GenY was to develop a better coding scheme for e-Coaching communications aimed at improving GenY eating habits. Segmentation of clinical conversation in the context of electronically delivered interventions into groups of MI behaviors, is traditionally performed manually by MI researchers, which significantly slows down their qualitative analysis. This paper is the first work to evaluate the empirical effectiveness of deep learning architectures in addressing \textit{the problem of discourse segmentation in the context of e-mail based behavioral interventions}.

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. We consider segmentation as a binary classification problem, in which each word or punctuation mark is annotated with a label of 1 or 0 to indicate whether it precedes a new segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning methods, such as Conditional Random Field (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (bidirectional RNN or BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} For an MLP model, training and testing instances were created based on a sliding window of $2n$ words or punctuation marks over each position in a given input sequence, such that each instance consists of a pair of the next $n$ words or punctuation marks and prior $n$ words or punctuation marks, including the current word or punctuation mark. Each sample is classified into either a ``new segment'' or a ``same segment'' based on whether there should be a segment break after the current word or not. In the case of CRF, BRNN and CRNN models, an email was taken as an input sequence, POS tags and word embeddings of each word or punctuation mark were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output of a model. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

\textbf{Convolutional Recurrent Neural Network}: CRNN is a deep neural network architecture,\cite{treviso2017sentence} shown in Figure~\ref{fig:crnn}, which consists of 5 layers: 1) input layer 2) embedding layer 3) convolution layer with max pooling 4) BRNN layer 5) fully connected layer with dropout and sigmoid output. 
E-coaching email exchanges are represented as a sequence of $m$ words and punctuations, which are fed into the input layer to produce a $m \times n_e$ matrix after fetching the pre-trained word vectors. When POS tags are used in combination with word embeddings, 10-dimensional POS vectors are concatenated with 300-dimensional word vectors to obtain new vectors $n_e = [n_w;n_p]$ of size 310. The primary purpose of convolution is to extract new features depending on the neighboring words. We used 1D convolution. In this layer, one filter is responsible for extraction of one feature. After applying $n_f$ different filters with zero-padding on both sides of the input text, $n_f$ useful features are produced in the convolution layer for each word. A max pooling operation was then performed over time to find the most significant features in a textual fragment. The bidirectional recurrent layer receives new features extracted in the convolution layer. RNN is usually used to capture long-range dependencies in a sequence of observations. Moreover, bidirectional RNNs are capable of capturing both past and future contexts through forward and backward traversals of a sequence. The purpose of the fully connected layer is to use the output of bidirectional RNN layer for classifying each word or punctuation into ``new segment'' and ``same segment'' classes. Since a fully connected layer has a greater number of parameters, they are more likely to excessively co-adapt with these parameters, which may cause over-fitting. To prevent this, we utilized dropout by randomly disregarding 50\% of the connections to the fully connected layer. After that, logistic sigmoid is applied to the output layer to provide the probability of whether the word or punctuation precedes a new segment or not. We experimentally determined the optimal parameters and found that the best performance is achieved with 5-fold cross-validation when filter length in convolutional layer is 7, number of filters is 100, max-pool size is 3, ReLU is used as a convolutional layer activation function, RNN layer activation is hyperbolic tangent and the number of recurrent units in RNN is 200. Adam\cite{kingma2014adam} with 50 epochs, the batch size of 32 and learning rate of 0.001 was used for optimization and the early stopping strategy was applied.\footnote{source code will be available at https://github.com/teanalab/xxxxx upon acceptance of this paper for publication}

@@

\subsection*{\textit{Evaluation metrics}}
We report standard metrics of precision, recall and F1-measure to evaluate the performance of the classifiers.\cite{aas1999text} Accuracy is not reported as a performance metric because it is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem, when highly unbalanced datasets are involved. The results are reported based on 5 fold cross-validation and weighted macro-averaging over the folds. We also estimate the area under the receiving operating characteristics (ROC) curve\cite{kumar2011receiver} (AUC) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets. \cite{hu2015kernelized}

@@

Table~\ref{tab:result_roc} shows the impact of individual features as well as their combination on the segmentation task. Although CRF provides the second highest classification results, it shows best AUC values, achieving AUC 0.966 when only lexical features are used and AUC 0.994 when a combination of lexical, punctuation and POS features are used. Influence of the punctuation and POS features is also consistent in AUC values, which increase by 2.9\%, 4.31\%, 6.63\% and 2.18\% for CRF, MLP, BRNN and CRNN methods, respectively, when all features are used together. Individually, POS and punctuation features also improve the performance of all classifiers except CRNN when word embeddings are used with POS features. CRF and MLP achieve their highest AUCs when all features are used together. On the other hand, BRNN and CRNN demonstrate the highest AUCs when word embeddings are combined with punctuation features. \\

@@

Punctuation mark and POS features resulted in significant improvement in the performance of machine learning and all deep learning methods. Especially, punctuation features have the higher individual impact on model performance compared to POS features. In almost all cases, every method performs better, when word embeddings are used in conjunction with punctuation and POS features. This results also indicate that segmentation performance might be improved by adding additional relevant features.

@@

\bibliographystyle{vancouver}
\bibliography{manuscript}

>>>>

\documentclass{amia}
\usepackage{graphicx}
\usepackage[labelfont=bf]{caption}
\usepackage[superscript,nomove]{cite}
\usepackage{color}
\usepackage{soul}
\usepackage{multirow}
\renewcommand*{\thefootnote}{\alph{footnote}}

@@

\section*{Introduction}
The emergence of e-Health technologies opened up new ways to deliver a variety of behavioral interventions to any demographic group of patients in any geographical location. Motivational interviewing (MI), an evidence-based communication technique to increase intrinsic motivation and self-efficacy for behavior change,\cite{miller2012motivational,miller2009toward} is one type of these interventions. MI sessions are generally aimed at eliciting ``change talk'', or statements of intrinsic motivation about patients' own desire, ability, reasons, need for and commitment to behavior change, which have been established by previous research\cite{apodaca2009mechanisms} as a reliable mediator of health behavior change. \hl{However, communication science approaches to understanding the efficacy of MI are inherently limited by traditional qualitative coding methods}.

@@

The goal of this study is to assess the effectiveness of deep learning methods for the task of automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of qualitative analysis of this type of clinical communications. \hl {Specifically, for this study, we utilized the data from MENU GenY (Making Effective Nutrition Choices for Generation Y) }\cite{alexander2017motivations}, a recent technology-based public health intervention that relies on personalized e-Coaching to encourage increased fruit and vegetable intake among young adults, aged 21--30. The goal of MENU GenY was to develop a better coding scheme for e-Coaching communications aimed at improving GenY eating habits. Segmentation of clinical conversation in the context of electronically delivered interventions into groups of MI behaviors, is traditionally performed manually by MI researchers, which significantly slows down their qualitative analysis. This paper is the first work to evaluate the empirical effectiveness of deep learning architectures in addressing \textit{the problem of discourse segmentation in the context of e-mail based behavioral interventions}.

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
\hl{The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. We consider segmentation as a binary classification problem, in which each word or punctuation mark is annotated with a label of 1 or 0 to indicate whether it precedes a new segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances}. In this study, we experimented with traditional machine learning methods, such as Conditional Random Field (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (bidirectional RNN or BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} For an MLP model, training and testing instances were created based on a sliding window of $2n$ words or punctuation marks over each position in a given input sequence, such that each instance consists of a pair of the next $n$ words or punctuation marks and prior $n$ words or punctuation marks, including the current word or punctuation mark. Each sample is classified into either a ``new segment'' or a ``same segment'' based on whether there should be a segment break after the current word or not. In the case of CRF, BRNN and CRNN models, an email was taken as an input sequence, POS tags and word embeddings of each word or punctuation mark were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output of a model. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

\textbf{Convolutional Recurrent Neural Network}: CRNN is a deep neural network architecture,\cite{treviso2017sentence} shown in Figure~\ref{fig:crnn}, \hl {which consists of 5 layers: 1) input layer 2) embedding layer 3) convolution layer with max pooling 4) BRNN layer 5) fully connected layer with dropout and sigmoid output. 
E-coaching email exchanges are represented as a sequence of $m$ words and punctuations, which are fed into the input layer to produce a $m \times n_e$ matrix after fetching the pre-trained word vectors. When POS tags are used in combination with word embeddings, 10-dimensional POS vectors are concatenated with 300-dimensional word vectors to obtain new vectors $n_e = [n_w;n_p]$ of size 310. The primary purpose of convolution is to extract new features depending on the neighboring words. We used 1D convolution. In this layer, one filter is responsible for extraction of one feature. After applying $n_f$ different filters with zero-padding on both sides of the input text, $n_f$ useful features are produced in the convolution layer for each word. A max pooling operation was then performed over time to find the most significant features in a textual fragment. The bidirectional recurrent layer receives new features extracted in the convolution layer. RNN is usually used to capture long-range dependencies in a sequence of observations. Moreover, bidirectional RNNs are capable of capturing both past and future contexts through forward and backward traversals of a sequence. The purpose of the fully connected layer is to use the output of bidirectional RNN layer for classifying each word or punctuation into ``new segment'' and ``same segment'' classes. Since a fully connected layer has a greater number of parameters, they are more likely to excessively co-adapt with these parameters, which may cause over-fitting. To prevent this, we utilized dropout by randomly disregarding 50\% of the connections to the fully connected layer. After that, logistic sigmoid is applied to the output layer to provide the probability of whether the word or punctuation precedes a new segment or not}. We experimentally determined the optimal parameters and found that the best performance is achieved with 5-fold cross-validation when filter length in convolutional layer is 7, number of filters is 100, max-pool size is 3, ReLU is used as a convolutional layer activation function, RNN layer activation is hyperbolic tangent and the number of recurrent units in RNN is 200. Adam\cite{kingma2014adam} with 50 epochs, the batch size of 32 and learning rate of 0.001 was used for optimization and the early stopping strategy was applied.\footnote{\hl {source code will be available at https://github.com/teanalab/xxxxx upon acceptance of this paper for publication}}

@@

\subsection*{\textit{Evaluation metrics}}
We report standard metrics of precision, recall and F1-measure to evaluate the performance of the classifiers.\cite{aas1999text} Accuracy is not reported as a performance metric because it is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem, when highly unbalanced datasets are involved. The results are reported based on 5 fold cross-validation and weighted macro-averaging over the folds. \hl{We also estimate the area under the receiving operating characteristics (ROC) curve} \cite{kumar2011receiver} \hl{(AUC) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets}. \cite{hu2015kernelized}

@@

Table~\ref{tab:result_roc} \hl {shows the impact of individual features as well as their combination on the segmentation task. Although CRF provides the second highest classification results, it shows best AUC values, achieving AUC 0.966 when only lexical features are used and AUC 0.994 when a combination of lexical, punctuation and POS features are used. Influence of the punctuation and POS features is also consistent in AUC values, which increase by 2.9\%, 4.31\%, 6.63\% and 2.18\% for CRF, MLP, BRNN and CRNN methods, respectively, when all features are used together. Individually, POS and punctuation features also improve the performance of all classifiers except CRNN when word embeddings are used with POS features. CRF and MLP achieve their highest AUCs when all features are used together. On the other hand, BRNN and CRNN demonstrate the highest AUCs when word embeddings are combined with punctuation features}. \\

@@

Punctuation mark and POS features resulted in significant improvement in the performance of machine learning and all deep learning methods. Especially, punctuation features have the higher individual impact on model performance compared to POS features. \hl {In almost all cases, every method performs better, when word embeddings are used in conjunction with punctuation and POS features}. This results also indicate that segmentation performance might be improved by adding additional relevant features.

@@

\bibliographystyle{vancouver}
\bibliography{manuscript_marked}
